---
title: "ElPiGraph.R"
author: '[Luca Albergante](mailto:luca.albergante@gmail.com)'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  md_document:
    toc: yes
    toc_depth: 2
  html_document:
    depth: 3
    keep_md: yes
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Description

This package provides an R implementation of the ElPiGraph algorithm. A self-contained description of the algorithm is available [here](https://github.com/auranic/Elastic-principal-graphs/blob/master/ElPiGraph_Methods.pdf)

A native MATLAB implementation of the algorithm (coded by Andrei Zinovyev) is also [available](https://github.com/auranic/Elastic-principal-graphs)

# Installation

To improve the performance of the algorithm, a number of functions have been implemented as C++ functions. To simplify the maintenance and updating of the package, these functions have been implemented in the package `distutils`, which needs to be installed separately. The following command will check the presence of the `devtools`, and install it if necessary, after that it will install the `distutils` package. A working internet connection is required.

```{r inst.dist, eval=FALSE}
if(!require("devtools")){
  install.packages("devtools")
}
devtools::install_github("Albluca/distutils")  
```

Once `distutils` has been installed, `ElPiGraph.R` can be installed by typing 

```{r inst.Elpi, eval=FALSE}
devtools::install_github("Albluca/ElPiGraph.R")
```

It is also possible to get the most recent developmental version (which will contains more feature, but is potentially more unstable) via:

```{r inst.Elpi.Dev, eval=FALSE}
devtools::install_github("Albluca/ElPiGraph.R", ref = "Development")
```

The package can then be loaded via the command

```{r load.Elpi}
library("ElPiGraph.R")
```

# Usage

Several guides are available to exemplify the behavior of ElPiGraph.R:

* [Basic usage](guides/base.md)
* [Derivation of graph substructures](guides/struct.md)
* [Derivation and visualization of pseudotime](guides/pseudo.md)
* [Usage of the trimming radius](guides/trim.md)
* [Advanced bootstrapping (Currently unavailable)](guides/boot.md)
* [Generation of synthetic data (Currently unavailable)](guides/synth.md)
* [Plotting (Currently unavailable)](guides/plotting.md)
