---
title: "ElPiGraph.R basics"
author: '[Luca Albergante](mailto:luca.albergante@gmail.com)'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  md_document:
    toc: yes
    toc_depth: 2
  html_document:
    depth: 3
    keep_md: yes
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Constructing a principal graph

The construction of a principal graph with a given topology is done via the specification of an appropriate initial conditions and of appropriate growth/shrink grammars. This can be done via the `computeElasticPrincipalGraph`.

Specific wrapping functions are also provided to build commonly encountered topologies (`computeElasticPrincipalCurve`, ``computeElasticPrincipalTree`, `computeElasticPrincipalCircle`), with minimal required inputs. In all of these function, it is necessary to specify a numeric matrix with the data points (`X`) and the number of nodes to of the principal graph (`NumNodes`). It is possible to control the behavior of the algorithm via a set of optional parameters. For example, it is possible to:

* modify the parameter controlling the elastic energy (`Mu` and `Lambda`)
* specify the number of processor to be used (`n.cores`)
* indicate if diagnostic plots should be produced (`drawAccuracyComplexity` and `drawEnergy`)
* indicate if the final graph should be used to plotted non the data (`drawPCAView`)
* specify if PCA should be performed on the data prior to principal graph fitting (`Do_PCA`) and if dimensionality  should be reduced (`ReduceDimension`)

# Examples of principal curves

The function `computeElasticPrincipalCurve` constructs a principal curve on the data. For example to construct a principal curve with 50 nodes on the example dataset `line_data`, it is sufficient to write

```{r load.Elpi}
library("ElPiGraph.R")
```

```{r}
CurveEPG <- computeElasticPrincipalCurve(X = curve_data, NumNodes = 60)
```

A principal tree can be constructed via the `computeElasticPrincipalTree` function. For example to construct a principal tree with 50 nodes on the example dataset `tree_data`, it is sufficient to write

```{r}
TreeEPG <- computeElasticPrincipalTree(X = tree_data, NumNodes = 60, Lambda = .03, Mu = .01)
```


Finally, a principal circle can be constructed via the `computeElasticPrincipalCircle` function. For example to construct a principal circle with 50 nodes on the example dataset `circe_data`, it is sufficient to write


```{r}
CircleEPG <- computeElasticPrincipalCircle(X = circle_data, NumNodes = 40)
```


All of these functions will return a list of length 1, with all the information associated with the graph.

# Using bootstrapping

All of the functions provided to build principal graphs allow a bootstrapped construction. To enable that it is sufficient to modify the parameters `nReps` and `ProbPoint`. `nReps` indicates the number of repetitions and `ProbPoint` indicates the probability to include a point in each of the repetition. When `nReps` is larger than 1, a final consensus principal graph will be constructed using the nodes of the graph derived in each repetition.

As an example, let us perform bootstrapping on the circle data. We will also prevent the plotting, for now.

```{r}
set.seed(42)
CircleEPG.Boot <- computeElasticPrincipalCircle(X = circle_data, NumNodes = 40, nReps = 50, ProbPoint = .6,
                                                drawAccuracyComplexity = FALSE, drawEnergy = FALSE, drawPCAView = FALSE)
```

`CircleEPG.Boot` will be a list with 51 elements: the 50 bootstrapped circles and the final consensus one.

# Plotting data with principal graphs

The `ElPiGraph.R` provides different functions to explore show how the principal graph approximate the data. The main function is `plotPG`. This function can be used to show how the principal graph fit the data in different ways.

To plot the principal tree previously constructed we can type

```{r}
PlotPG(X = tree_data, TargetPG = TreeEPG[[1]], Main = "A tree")
```

The main plot reports different features including the percentage of variance explained relative to the nodes of the principal graph (PG var), the percentage of variance explained relative to the data points (Data var), the fraction of variance of the data explained by the nodes of the principal graph (FVE) and the fraction of variance of the data explained by the projection of the points on the the principal graph (FVEP). In this example the nodes of the principal graph have been used to compute PCA and rotate the space (the Do_PCA parameter is TRUE be default), this can be seen by the "EpG PC" label of the axes.

To include additional dimension in the plot it is sufficient to specify them with the DimToPlot parameter, e.g.,

```{r}
PlotPG(X = tree_data, TargetPG = TreeEPG[[1]], Main = "A tree", DimToPlot = 1:3)
```

We can also visualize the results of the bootstrapped construction by using the `BootPG` parameter:

```{r}
PlotPG(X = circle_data, TargetPG = CircleEPG.Boot[[length(CircleEPG.Boot)]],
       BootPG = CircleEPG.Boot[1:(length(CircleEPG.Boot)-1)],
       Main = "A bootstrapped circle", DimToPlot = 1:2)
```