---
title: "Usig pseudotime"
author: '[Luca Albergante](mailto:luca.albergante@gmail.com)'
date: "11 janvier 2018"
output:
  md_document:
    toc: yes
    toc_depth: 2
  html_document:
    depth: 3
    keep_md: yes
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The ElPiGraph package contains a number of functions to derive the pseudotime associated with each point. This is particularly relevant in biological contexts. 

# Setup

As a first step, we will construct a tree structure on the sample data

```{r}
library(ElPiGraph.R)

TreeEPG <- computeElasticPrincipalTree(X = tree_data, NumNodes = 50,
                                       drawAccuracyComplexity = FALSE, drawEnergy = FALSE)
```

and visualize the node labels:

```{r}
PlotPG(X = tree_data, TargetPG = TreeEPG[[1]],
       NodeLabels = 1:nrow(TreeEPG[[1]]$NodePositions),
       LabMult = 3, PointSize = NA, p.alpha = .1)
```


# Getting the substructure of interest

In this example, we will look at all the path originating in 2 and extending to all the other leafs of the tree. Hence, we assume that 2 is the starting points for the pseudotime.

We will begin by computing all of the paths between leaves.

```{r}
Tree_Graph <- ConstructGraph(TreeEPG[[1]])
Tree_e2e <- GetSubGraph(Net = Tree_Graph, Structure = 'end2end')
```

Since the paths derived are unique, the node 2 could be either at the beginning or at the end of the path considered. therefore, we select all the path starting or ending in 2 

```{r}
Root <- 2
SelPaths <- Tree_e2e[sapply(Tree_e2e, function(x){any(x[c(1, length(x))] == Root)})]
```

and reverse the paths ending in 2

```{r}
SelPaths <- lapply(SelPaths, function(x){
  if(x[1] == Root){
    return(x)
  } else {
    return(rev(x))
  }
})

```

At this point, we can look at `SelPaths` to make sure that the results are compatible with our expectations

```{r}
SelPaths
```

# Getting the supporting structures

To optimize the computation, certain structures used to compute the pseudotime are computed externally. We begin by computing a Partition structure, which contains information on the projection of points on the nodes

```{r}
PartStruct <- PartitionData(X = tree_data, NodePositions = TreeEPG[[1]]$NodePositions)
```

Then, we will compute a projection structure, which contains information relative to the projection of points on the edge of the graph

```{r}
ProjStruct <- project_point_onto_graph(X = tree_data,
                                       NodePositions = TreeEPG[[1]]$NodePositions,
                                       Edges = TreeEPG[[1]]$Edges$Edges,
                                       Partition = PartStruct$Partition)
```

# Computnig pseudotime

We are now able to obtain the pseudotime, via the `getPseudotime` function. We will use `lapply` to compute all of the pseudotime at once. Moreover, the functions requires nodes to be passed as a vector of strings, so we will need to use the `names` function to convert the sequence of vertices.

```{r}
AllPt <- lapply(SelPaths, function(x){
  getPseudotime(ProjStruct = ProjStruct, NodeSeq = names(x))
})
```

At this point, we can merge the pseudotime projections. This is possible because points are uniquely projected on the graph and all the paths selected have a common root

```{r}
PointsPT <- apply(sapply(AllPt, "[[", "Pt"), 1, function(x){unique(x[!is.na(x)])})
```

When can then visualize the pseudotime on the points via the `PlotPG` function 

```{r}
PlotPG(X = tree_data, TargetPG = TreeEPG[[1]], GroupsLab = PointsPT)
```

# Exploring





