---
title: "Use structure to redefine dimensionality"
author: '[Luca Albergante](mailto:luca.albergante@gmail.com)'
date: "11 janvier 2018"
output:
  md_document:
    toc: yes
    toc_depth: 2
  html_document:
    depth: 3
    keep_md: yes
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Once a tree structure has been defined, it is possible to use it to redefine the dimensionality of the data in such way to obtain a structure dependend dimension

# Setup

At first we will create a tree ElPiGraph.

```{r}
library(ElPiGraph.R)
library(igraph)

TreeEPG <- computeElasticPrincipalTree(X = tree_data, NumNodes = 50,
                                       drawAccuracyComplexity = FALSE, drawEnergy = FALSE)
```

We then need to select a node that will act as a root for the analysis. To do so, we need to visualize the labels of the nodes. To simplify the operation, we will only visualize the id of nodes with degree 1 (i.e., the leaves).

```{r}
Net <- ConstructGraph(PrintGraph = TreeEPG[[1]])
Nodelab <- 1:vcount(Net)
Nodelab[degree(Net) > 1] <-NA

PlotPG(X = tree_data, TargetPG = TreeEPG[[1]], NodeLabels = Nodelab, LabMult = 8)
```

For the following example we will use 2 as a the root

```{r}
BrcStruct <- BranchingDimension(X = tree_data, PrintGraph = TreeEPG[[1]], Start = "2")
```

The `BrcStruct$DimMatrix` will now contain the points re-dimensionalised. Each column of the matrix contain a branch dimension between 0 and1

```{r}
BrcStruct$DimMatrix[is.na(BrcStruct$DimMatrix)] <- 0
pheatmap::pheatmap(BrcStruct$DimMatrix)
```



